Efficient pedigree recording for fast population genetics simulation

### 2018-01-23 forwards_paper.tex
added note about sortedness
minor changes re jda comments

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\paragraph{Essential steps in tree sequence recording}
To record a tree sequence, including mutations,
a simulator must record:
\begin{itemize}
    \item the birth time of each individual in the Node Table,
    \item the genetic segments each individual inherits from her parents in the Edge Table, and
    \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
        adding also to the Site Table if the site is previously unmutated.
\end{itemize}
Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
Before simplification, time must be translated to ``time ago'' if it is not already.
which was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.
Furthermore, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.
% Finally, the first step in simplification is to \emph{sort} the tables
% according to several criteria, for algorithmic efficiency,
% as described in the documentation.

@@

\begin{figure}
    \begin{center}
        \includegraphics{simplify-state-diagram}
    \end{center}
    \caption{
        A depiction of the state of the simplification algorithm
        as it moves up through the embellished pedigree
        in the example of Figure~\ref{fig:method_diagram}A.
        Following the ``paint pot'' description in the text,
        we begin by coloring J and K's genomes in red and blue respectively,
        then trace how these colors were inherited back up through the pedigree
        until they coalesce.
        To aid in this,
        the smaller colored chromosomes on either side of each solid arrow
        show the bits inherited from each of the two parental chromosomes,
        with genomic position 0.0 on the bottom and 1.0 at the top.
        Each time a red and a blue segment overlap, a coalescence occurs, two edges are output,
        and we stop propagating that segment.
        For instance, both J and K inherit from H between 0.5 and 0.9,
        which resulted in the first two edges of the simplified table of Figure~\ref{fig:method_diagram}C.
        Later, both inherit from E between 0.2 and 0.5,
        along the paths J-H-G-E and K-I-E respectively,
        resulting in the next two edges.
        \label{fig:simplify_state}
    }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Gil McVean, Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
the NIH (R01GM115564) to KRT.

@@

Here we provide a concrete implementation of the simplification algorithm.
The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
These output node IDs correspond to colors in the ``paint pot'' analogy.
The per-interval segments are assigned to input IDs via the $\Al$ mapping,
which is the main state required by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node IDs mapped to input
node ID $u$ over a set of intervals.
For each input node, we need to process the ancestral segments
to find overlaps (and hence coalescences), and to update $\Al$.
We do this using a $\priorityq{}$ type, which maintains a list of ancestral
segments sorted by left coordinate. For a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
removes and returns the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.

>>>>

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The leftmost panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The remaining panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\paragraph{Essential steps in tree sequence recording}
To record a tree sequence, including mutations,
a simulator must record:
\begin{itemize}
    \item the birth time of each individual in the Node Table,
    \item the genetic segments each individual inherits from its parents in the Edge Table, and
    \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
        adding also to the Site Table if the site is previously unmutated.
\end{itemize}
Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
Before simplification, time must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
Furthermore, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.
% Finally, the first step in simplification is to \emph{sort} the tables
% according to several criteria, for algorithmic efficiency,
% as described in the documentation.

@@

\begin{figure}
    \begin{center}
        \includegraphics{simplify-state-diagram}
    \end{center}
    \caption{
        A depiction of each state of the simplification algorithm
        as it moves up through the embellished pedigree
        in the example of Figure~\ref{fig:method_diagram}A.
        Following the ``paint pot'' description in the text,
        we begin by coloring J and K's genomes in red and blue respectively,
        then trace how these colors were inherited back up through the pedigree
        until they coalesce.
        To aid in this,
        the smaller colored chromosomes on either side of each solid arrow
        show the bits inherited from each of the two parental chromosomes,
        with genomic position 0.0 on the bottom and 1.0 at the top.
        Each time a red and a blue segment overlap, a coalescence occurs, two edges are output,
        and we stop propagating that segment.
        For instance, both J and K inherit from H between 0.5 and 0.9,
        which resulted in the first two edges of the simplified table of Figure~\ref{fig:method_diagram}C.
        Later, both inherit from E between 0.2 and 0.5,
        along the paths J-H-G-E and K-I-E respectively,
        resulting in the next two edges.
        \label{fig:simplify_state}
    }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Gil McVean, Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
the NIH (R01GM115564) to KRT;
and the USF\&WS to H.\ Bradley Shaffer.

@@

Here we provide a concrete implementation of the simplification algorithm.
The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
These output node IDs correspond to colors in the ``paint pot'' analogy.
The per-interval segments are assigned to input IDs via the $\Al$ mapping,
which is the main state required by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node IDs mapped to input
node ID $u$ over a set of intervals.
For each input node, we need to process the ancestral segments
to find overlaps (and hence coalescences), and to update $\Al$.
We do this using a $\priorityq{}$ type, which maintains a list of ancestral
segments sorted by left coordinate. For a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
removes and returns the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.
Finally, note that to keep the exposition simple,
the algorithm described here requires that the input node table
is sorted by time since the parent's birth;
but the version implemented in \msprime{} makes different sortedness assumptions
(described in the documentation).
